The Weakly Nonlinear Magnetorotational Instability

### 2016-09-02 paper/widegap_mri_paper.tex

\section{Weakly Nonlinear Perturbation Analysis}\label{sec:wnl_analysis}
\label{sec:Perturbation}
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_coeffs_satampscaling.png}
\caption{Critical parameters $\reym_c$ and $k_c$, and coefficients of the Ginzburg-Landau equation (Equation \ref{eq:gle}) as a function of $\Pm$. Note the very weak dependence of the linear ($b$) and diffusive ($h$) coefficients on $\Pm$. The saturation amplitude $\alpha_{sat} = \sqrt{b/c}$ of the standard MRI system has a power law dependence on $\Pm$ which we measure to be $\alpha_{sat} \sim \Pm^{0.777}$. This scaling is driven by the $\Pm$ dependence of the nonlinear coefficient $c$.}\label{fig:coefficients}
\end{figure}

@@

\subsection{Helical MRI}
\label{sec:HMRI}
When $\xi$ in Equation \ref{eq:unperturbed_matrix_eqns} is not equal to zero, the helical MRI arises. We examine a single fiducial helical MRI case, for the parameters listed in Table \ref{table:parameters}. At the conclusion of the weakly nonlinear analysis, we find that the coefficients of Equation \ref{eq:gle} are $b = $, $d = $, and $c = $, i.e. complex. The marginal helical MRI is thus described by a complex GLE. This difference in character between the amplitude equations that modulate the weakly nonlinear standard and helical MRI is a consequence of the same property that makes the helical MRI an overstability. With the introduction of an azimuthal component, the background magnetic field acquires a handedness that is not present in a purely axial field.

@@

\section{Acknowledgements}
JSO acknowledges support from NASA grant 15-LWS15\_2-0063. NSF. The authors thank Mordecai Mac Low, Jeremy Goodman, John Krommes, Geoff Vasil, etc for useful discussion.

>>>>

\section{Weakly Nonlinear Perturbation Analysis}\label{sec:wnl_analysis}
\label{sec:Perturbation}
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_coeffs_satampscaling_fivepoints.png}
\caption{Critical parameters $\reym_c$ and $k_c$, and coefficients of the Ginzburg-Landau equation (Equation \ref{eq:gle}) as a function of $\Pm$. Note the very weak dependence of the linear ($b$) and diffusive ($h$) coefficients on $\Pm$. The saturation amplitude $\alpha_{sat} = \sqrt{b/c}$ of the standard MRI system has a power law dependence on $\Pm$ which we measure to be $\alpha_{sat} \sim \Pm^{0.777}$. This scaling is driven by the $\Pm$ dependence of the nonlinear coefficient $c$.}\label{fig:coefficients}
\end{figure}

@@

\subsection{Helical MRI}
\label{sec:HMRI}
When $\xi$ in Equation \ref{eq:unperturbed_matrix_eqns} is not equal to zero, the helical MRI arises. We examine a single fiducial helical MRI case, for the parameters used by \citei{Hollerbach:2005tr}, listed in Table \ref{table:parameters}. The helical MRI is an overstability, so the ansatz linear eigenvector we consider (Equation \ref{eq:ansatz}) is characterized by a complex temporal eigenvalue $\sigma$. For our fiducial parameters, the marginal mode has a frequency $\omega = 0.153$. This means that the helical MRI modes are traveling waves, moving in the $z$ direction with a phase velocity $\omega/k_c$ = .

At the conclusion of the weakly nonlinear analysis, we find that the coefficients of Equation \ref{eq:gle} are complex. The marginal helical MRI is thus described by a complex GLE. This difference in character between the amplitude equations that modulate the weakly nonlinear standard and helical MRI is a consequence of the same property that makes the helical MRI an overstability. With the introduction of an azimuthal component, the background magnetic field acquires a handedness that is not present in a purely axial field. The helical MRI eigenvectors are therefore free to be out of phase with one another. In our perturbation series, the helical MRI modes interact within and between orders with modes which carry different phases, leading to complex GLE coefficients.

The phase dynamics of the complex GLE are well-studied in a variety of systems, and depend on the values of the GLE coefficients. The complex GLE may be unstable to traveling wave instabilities such as the Benjamin-Feir instability, a generalization of the Ekhaus instability which acts on stationary waves, and to which the standard MRI system may be unstable. The complex GLE can also admit spatiotemporal chaos, and various classes of coherent structures (\citei{Aranson:2002}).
%Because the saturation amplitude of the weakly nonlinear helical MRI is governed by a complex GLE, the saturated state of the helical MRI may be unstable to 
Although a detailed description of the phase dynamics in the helical MRI is beyond the scope of this work, we note that such long-wavelength, long-timescale behavior may be observed in liquid metal helical MRI experiments.

@@

The theory presented here may be extended in several ways.

\section{Acknowledgements}
S.E.C. was supported by a National Science Foundation Graduate Research Fellowship under grant No. DGE-11-44155. J.S.O. acknowledges support from NASA grant 15-LWS15\_2-0063. The authors thank Mordecai Mac Low, Jeremy Goodman, John Krommes, Geoff Vasil, Ellen Zweibel, etc for useful discussion.
